---
title: "Drug-tolerant melanoma RNAseq analysis"
output: html_notebook
author: Darren Tyson & Philip Stauffer
date: 2023-08-29
---
# Analysis of RNA-seq data
Four BRAF-mutant melanoma cell lines (A375, SKMEL28, SKMEL5 and WM88) in standard culture medium (naive) or treated with BRAFi (8µM PLX4720) for 8 days (tolerant). Illumina Novaseq paired end reads.


```{r Setup, message=FALSE, warning=FALSE}
library(DESeq2)
library(EnhancedVolcano)
library(org.Hs.eg.db)
library(ggplot2)

OVERWRITE <- FALSE
OUTPUT_DIR <- file.path("~/P2RX7_ms_output")
if(!dir.exists(OUTPUT_DIR)) dir.create(OUTPUT_DIR)
```


```{r}
d <- read.csv("../data/STARquantMode_genecounts_all.csv", row.names = 1)
genes <- rownames(d)

samples <- colnames(d)
cell_lines <- factor(sapply(strsplit(samples, "_"), "[[", 1))
conds <- factor(ifelse(sapply(strsplit(samples, "_"), "[[", 2)=="8day","tolerant","naive"))
reps <- factor(gsub("rep","",sapply(strsplit(samples, "_"), "[[", 3)))
coldat <- data.frame(cell_line=cell_lines,condition=conds,rep=reps)
coldat$group <- factor(paste(coldat$cell_line,coldat$condition,sep="_"))

lowgenes <- rownames(d[apply(d,1,function(x) all(x<10)),])
```

### Make DESeq dataset
```{r}
dds <- DESeqDataSetFromMatrix(countData = d,
                              colData = coldat,
                              design = ~ cell_line + condition)

```

### Set reference condition
Using drug-tolerant A375 cells as reference.

Maintaining low-expressed genes reduces the impact of cell line-specific differences in gene expression. 
```{r}
dds$cell_line <- relevel(dds$cell_line, ref = "A375")
dds$condition <- relevel(dds$condition, ref = "tolerant")

dds <- DESeq(dds)
```

### Normalized counts
```{r}
dds <- estimateSizeFactors(dds)
normalized_counts <- counts(dds, normalized=TRUE)
```

### Variance stabilization for PCA
```{r}
res <- results(dds)
vsd <- vst(dds, blind=FALSE)
```

### % of genes that are both `is.na(padj)` and in `lowgenes`
Missing values for padj are highly enriched in low-expressed genes.
```{r}
cat(paste(signif(length(which(rownames(res)[which(is.na(res$padj))] %in% lowgenes))/length(rownames(res)[which(is.na(res$padj))])*100,4),"% of genes with padj == NA are in lowgenes"))
```

### PCA of variance-stabilized data
```{r}
pcaData <- plotPCA(vsd, intgroup=c("condition", "cell_line"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
ggplot(pcaData, aes(PC1, PC2, color=condition, shape=cell_line)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()
```


### Specifically examine tolerant vs naive
Ignoring cell line-specific differences. 
```{r}
res_tolerant <- results(dds, contrast=c("condition","tolerant","naive"))
res_tolerant <- res_tolerant[order(res_tolerant$padj),]
res_tolerant <- res_tolerant[!is.na(res_tolerant$padj),]
summary(res_tolerant)
```

## Volcano plot of differential gene expression
```{r fig.height=5, fig.width=6, warning=FALSE}
ggp <- EnhancedVolcano(res_tolerant,
    lab = rownames(res_tolerant),
    x = 'log2FoldChange',
    y = 'pvalue',
    pCutoff = 1e-03,
    pCutoffCol = "padj",
    title = "Drug-tolerant vs naive",
    subtitle = bquote(italic("all samples, all genes")),
    axisLabSize=9,
    labSize=1.5, legendLabSize=7, legendIconSize=1, pointSize=1,
    titleLabSize=12, subtitleLabSize=10, captionLabSize=10,
    shape = 20,
    col = c("grey30", "grey60", "orange", "red2"),
    max.overlaps = 20,
    borderWidth = 0.6,
    encircleSize = 1
    )
ggp
```

```{r}
mylabels <- c("ABCB5","MERTK","NGFR","AXL","EGFR","TYRP1","DCT","EDNRB","PPARGC1A",
              "TRPM1","TRPM8","P2RX4","P2RX7","CACNA1C","PANX2","CACNA1D",
              "TRPC4","NFATC4",
              "SLC24A2","SPRY4","EREG","MITF",
              "ETV5","ETV4","SPRY1",
              "CCND2","CXCL12")
```



```{r fig.height=5, fig.width=4, warning=FALSE}
ggp <- EnhancedVolcano(res_tolerant,
    lab = rownames(res_tolerant),
    selectLab = mylabels,
    x = 'log2FoldChange',
    y = 'pvalue',
    xlim=c(-6,9),
    ylim=c(0,75),    
    pCutoff = 1e-03,
    pCutoffCol = "padj",
    title = "Drug-tolerant vs naive",
    subtitle = bquote(italic("all samples")),
    axisLabSize=9,
    labSize=1.25, legendLabSize=7, legendIconSize=1, pointSize=1,
    titleLabSize=12, subtitleLabSize=10, captionLabSize=10,
    shape = 20,
    col = c("grey30", "grey60", "orange", "red2"),
    drawConnectors = TRUE,
    widthConnectors = .5,
    typeConnectors = "closed",
    endsConnectors = "last",
    boxedLabels = TRUE,
    max.overlaps = 20,
    borderWidth = 0.6,
    encircleSize = 1
    )
ggp

fn <- file.path(OUTPUT_DIR,"Volcano_avg_taller.pdf")
if(OVERWRITE | !file.exists(fn))
  ggsave(file=fn, device="pdf", width=4, height=5, units="in")
```

### Gene enrichment ontology analysis
Using Benjamini-Hochberg-adjusted p-value of 0.05 and 1.5-fold change of expression as cutoffs.
```{r}
upreg_genes <- subset(res_tolerant, padj<0.05 & log2FoldChange>0.5849625)
downreg_genes <-subset(res_tolerant, padj<0.05 & log2FoldChange<(-0.5849625))

geneList_up <- as.vector(upreg_genes$log2FoldChange)
names(geneList_up) <- rownames(upreg_genes)
geneList_down <- as.vector(downreg_genes$log2FoldChange)
names(geneList_down) <- rownames(downreg_genes)

genes_up <- as.vector(rownames(upreg_genes))
genes_down <- as.vector(rownames(downreg_genes))
```

### Using `clusterProfiler` to perform gene ontology enrichment analysis
https://bioconductor.org/packages/release/bioc/vignettes/clusterProfiler/inst/doc/clusterProfiler.html

```{r}
ggo_up <- clusterProfiler::groupGO(gene  = genes_up,
                                OrgDb    = org.Hs.eg.db,
                                keyType  = "SYMBOL",
                                ont      = "MF",
                                level    = 3,
                                readable = TRUE)
ggo_up <- as.data.frame(ggo_up)
ggo_up <- ggo_up[order(-ggo_up$Count),] 
ego_genesUp <- clusterProfiler::enrichGO(gene  = genes_up,
                                 OrgDb         = org.Hs.eg.db,
                                 keyType  = "SYMBOL",
                                 ont           = "MF",
                                 pAdjustMethod = "BH",
                                 pvalueCutoff  = 0.05,
                                 qvalueCutoff  = 0.05, 
                                 readable      = TRUE)
```

```{r}
clusterProfiler::dotplot(ego_genesUp) + 
  ggtitle("GO Over-representation Upregulated Genes") +
  labs(x="Gene Ratio", y="GO Terms") +
  theme(legend.text = element_text(size = 8),
        plot.title = element_text(size = 12, hjust = 0.5, face = "bold"), 
        axis.text=element_text(size=8),
        legend.title = element_text(size=8,face="bold"), 
        axis.title=element_text(size=9, face="bold"))

fn <- file.path(OUTPUT_DIR,"GO_upreg.pdf")
if(OVERWRITE | !file.exists(fn))
  ggsave(file=fn, device="pdf", width=6, height=4, units="in")
```

```{r}
ggo_down <- clusterProfiler::groupGO(gene  = genes_down,
                                OrgDb    = org.Hs.eg.db,
                                keyType  = "SYMBOL",
                                ont      = "MF",
                                level    = 3,
                                readable = TRUE)
ggo_down <- as.data.frame(ggo_down)
ggo_down <- ggo_down[order(-ggo_down$Count),] 
ego_genesDown <- clusterProfiler::enrichGO(gene  = genes_down,
                                 OrgDb         = org.Hs.eg.db,
                                 keyType       = "SYMBOL",
                                 ont           = "MF",
                                 pAdjustMethod = "BH",
                                 pvalueCutoff  = 0.05,
                                 qvalueCutoff  = 0.05, 
                                 readable      = TRUE)
```

```{r}
clusterProfiler::dotplot(ego_genesDown) + 
  ggtitle("GO Over-representation Downregulated Genes") +
  labs(x="Gene Ratio", y="GO Terms") +
  theme(legend.text = element_text(size = 8),
        plot.title = element_text(size = 12, hjust = 0.5, face = "bold"), 
        axis.text=element_text(size=8),
        legend.title = element_text(size=8,face="bold"), 
        axis.title=element_text(size=9, face="bold"))
```



```{r}
upgenes_filtered <- as.data.frame(upreg_genes[order(upreg_genes$log2FoldChange,
                                                    decreasing=TRUE),])

upgenes_filtered <- subset(upgenes_filtered, log2FoldChange>1.5 & 
                             padj < 1e-4 & baseMean > 200)

### Exclude LINC (long noncoding RNAs) and LOC (unannotated) genes
upgenes_filtered <- subset(upgenes_filtered, !grepl("^L[OI][CN]",rownames(upgenes_filtered)))

upgenes_filtered
```


Calcium ion transport gene ontologies:  
https://www.informatics.jax.org/vocab/gene_ontology/GO:0006816

calcium ion transmembrane transport  
GO:0070588  

Amino acid transporter 
GO:0006865  

monoatomic cation channel activity
GO:0005261

```{r}
go_enriched2 <- strsplit(ego_genesUp@result["GO:0005261","geneID"],"/")
names(go_enriched2) <- "GO:0005261"

go_labels <- unlist(go_enriched2[[1]])
x <- res_tolerant[go_labels,c("log2FoldChange","padj")]
x <- x[order(x$log2FoldChange, decreasing=TRUE),]
go_labels <- rownames(x)

go_labels2 <- c("KCND3","KCNQ3","KCNB1","P2RX7","KCNJ13","P2RX1","TMEM63C","CACNA1D","CACNA1C")
```

```{r fig.height=5, fig.width=6, warning=FALSE}
ggp <- EnhancedVolcano(res_tolerant,
    lab = rownames(res_tolerant),
    selectLab = go_labels,
    title = "Genes associated with monoatomic cation channel activity",
    subtitle = bquote(italic(GO:0005261)),
    caption = paste0("total = ", nrow(res_tolerant), " genes"),
    x = 'log2FoldChange',
    y = 'pvalue',
    pCutoff = 1e-02,
    pCutoffCol = "padj",
    xlim=c(-1,6),
    ylim=c(0,60),
    axisLabSize=9,
    labSize=1.25, legendLabSize=7, legendIconSize=1, pointSize=1,
    titleLabSize=12, subtitleLabSize=10, captionLabSize=10,
    shape = 20,
    col = c("grey30", "grey60", "orange", "red2"),
    drawConnectors = TRUE,
    widthConnectors = .5,
    typeConnectors = "closed",
    endsConnectors = "last",
    boxedLabels = TRUE,
    max.overlaps = 20,
    borderWidth = 0.6,
    encircleSize = 1
    )
ggp

fn <- file.path(OUTPUT_DIR,"Volcano_avg_GO-5261.pdf")
if(OVERWRITE | !file.exists(fn))
  ggsave(file=fn, device="pdf", width=6, height=4, units="in")
```

### Plotting function for gene expression levels across cell line data (drug-tolerant vs naive)
```{r}
plot_naive_v_tol <- function(gene, yl=c(-2,18)){
  gene <- gene[1] # use only first item
  x <- cbind(coldat,gene=as.integer(normalized_counts[gene,]))
  g <- ggplot(x, aes(cell_line, log2(gene), color=condition, shape=condition)) +
    geom_jitter(width = .075, size=2) + 
    theme_classic() + 
    theme(legend.position = "top", legend.direction = "horizontal") + 
    theme(legend.title = element_text(size=0), legend.text = element_text(size=8)) + 
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
    labs(y="log2(normalized counts)", x="", title=gene) + 
    scale_y_continuous(limits=yl, breaks=seq(yl[1],yl[2],2))
  g
}
```


```{r fig.height=3, fig.width=1.5}
plot_naive_v_tol("ATP1A1", yl=c(14,19))
plot_naive_v_tol("P2RX7")
```


```{r}
downreg_genes["ITPR2",]
```



```{r fig.height=3, fig.width=1.5}
plot_naive_v_tol("ITPR2", yl=c(8,13))
plot_naive_v_tol("ITPR3",yl=c(11,15))
```


```{r}
downreg_genes["CALB2",]
```

```{r fig.height=3, fig.width=1.5}
p <- list(plot_naive_v_tol("CACNA1C",yl=c(-2,8)),
          plot_naive_v_tol("CACNA1D",yl=c(2,10)),
          plot_naive_v_tol("P2RX4",yl=c(4,14)),
          plot_naive_v_tol("P2RX1", yl=c(-2,8)),
          plot_naive_v_tol("P2RX7", yl=c(2,16)),
          plot_naive_v_tol("ITPR2",yl=c(8,13)),
          plot_naive_v_tol("CALB2", yl=c(-2,12)),
          plot_naive_v_tol("ATP2B2", yl=c(-2,8)))
p
```

