---
title: "Calcium activation of pERK"
output: html_notebook
author: Darren Tyson and Philip Stauffer
date: August 10, 2023
---

# Simplifying code 
Based on code in `./Ca2Mobilization/20230731_ionomycinCa2+mobilization/20230731_analysis_PES_v02.Rmd`

```{r Setup, message=FALSE, warning=FALSE}
library(diprate)
library(ggplot2)
library(viridis)
library(tidyr)
library(forcats)
library(dplyr)

SAVEPLOTS <- FALSE
```

### Function to quantify ppERK using both proportion of positive cells and the avg intensity
Using both proportion of cells staining positively for ppERK and the integrated values of the positive cells.  

```{r}
ppERKint <- function(dat, signif=4) {
    cell_count <- nrow(dat)
    pos_cells <- nrow(dat[dat$positive.w2==1,])
    val <- pos_cells/cell_count*sum(dat[dat$positive.w2==1,]$avg.pERK)
    val <- signif(val,signif)
    return(val)
}
```


## Data on DropBox
```{r}
dat_filepath <- "https://www.dropbox.com/scl/fi/y6d16tdkbgq77llv644aj/20230731_singlecell_all_iono.csv?rlkey=mfc5bm8xedpysoixilfuqrut5&dl=1"
d <- read.csv(dat_filepath)
```

### Perform some QC
```{r}
d <- d[d$total.area < 4500 & d$total.area > 100 & d$w2.stain.integr.intens <5e6,]
```

### Add new column: `avg.pERK`
```{r}
d$avg.pERK <- log(d$w2.stain.integr.intens/d$w2.stain.area)
d$avg.pERK <- d$avg.pERK - min(d$avg.pERK, na.rm=TRUE) # subtract lowest pERK level
d$avg.pERK[is.na(d$avg.pERK)] <- -0.5 # replace missing values with -0.5
```

### Add Ionomycin treatment times
```{r}
d$iono.tx.time <- as.integer(d$drug2)
d$iono.tx.time[is.na(d$iono.tx.time)] <- 0
```


### Filtering steps previously applied to second data set
2023-08-15 data have already been processed with the filtering steps above; data will be combined into single file
```{r}
d$expt.date <- ifelse(grepl("142319",d$plate.id),"20230729","20230815")
```


### Annotation
Controls: 
'drug1 == "control"' Drug naive cells in FBS containing media to compare the effects between FBS and no FBS in the media
`drug1 == "" & drug2 == ""` are for (1° and 2°) antibody controls

Separate controls and ionomycin-treated
```{r}
d$drug2[is.na(d$drug2)] <- ""
ab_ctrl <- d[d$drug1=="" & d$drug2=="", ]

# undrugged media was added to these wells; PLX4720 concentration was approximately 4 uM, down from 8 uM, for 20 minutes before fixation.
persistent <- d[d$drug1== "persistent", ] 

nFBS <- d[d$drug1=="control", ] # naive & FBS

ctrl_wells <- c(unique(ab_ctrl$well), unique(persistent$well), unique(nFBS$well))

tx <- d[!d$well %in% ctrl_wells, ]
tx$drug1.conc <- factor(signif(tx$drug1.conc, digits= 2))
tx$iono.tx.time <- factor(tx$iono.tx.time)

ctrls <- d[d$well %in% ctrl_wells, ]
ctrls <- ctrls[!ctrls$well %in% unique(ab_ctrl$well), ]
```

### Exclude some control wells
Philip's note: outlier wells may include C01, C02, D01, D02 F01, G01, H01
```{r}
# remove wells from 20230729 data
bad_wells_729 <- c("C01", "C02", "D01", "D02", "F01", "G01", "H01", "H02")

tx <- tx[!(tx$well %in% bad_wells_729 & tx$expt.date=="20230729"),]

#generating individual groups
day3 <- tx[tx$BRAFi.tx.time == "3 days", ]

min60 <- tx[tx$BRAFi.tx.time== "60 min",]

day3_grouped <- day3[!(day3$well %in% bad_wells_729 & day3$expt.date=="20230729"),] %>%
    group_by(expt.date, drug1, iono.tx.time, drug1.conc, positive.w2) %>%
    summarise(num=n()) 

day3_grouped$positive.w2 <- c("negative","positive")[match(day3_grouped$positive.w2,c(0,1))]

pct.pos <- day3_grouped %>%
    spread(positive.w2, num) %>%
    mutate(total= negative+positive) %>%
    summarise(propos= positive/total) %>% 
    ungroup()

pct.pos$total <- as.vector(day3_grouped[day3_grouped$positive.w2=="negative",]$num) + 
    as.vector(day3_grouped[day3_grouped$positive.w2=="positive",]$num)
```


### Plot the full distribution of ppERK cell staining for each condition
  
  
```{r}
library(ggridges)

dat <- tibble(tx[tx$expt.date=="20230729",c("well","positive.w2","BRAFi.tx.time","drug1.conc","iono.tx.time","avg.pERK")])
dat <- dat[!dat$well %in% bad_wells_729,]
dat$condition <- paste(dat$BRAFi.tx.time, dat$drug1.conc, sep="_")
dat <- dat[order(dat$BRAFi.tx.time,dat$drug1.conc),]
rownames(dat) <- NULL

drug_conc <- sort(unique(dat$drug1.conc))


plotRidges <- function(tib, dc, tit="Drug-tolerant cells") {
    tx.times <- unique(tib$iono.tx.time)
    n_vals <- do.call(rbind, lapply(tx.times, function(tt) data.frame(
                                        tx.time=tt,
                                        pos=length(which(tib[tib$iono.tx.time==tt,]$positive.w2==1)),
                                        total=nrow(tib[tib$iono.tx.time==tt,]))
                                    )
                      )
    n_vals$propos <- signif(n_vals$pos/n_vals$total*100, 3)
    
    ppERK_vals <- sapply(sort(unique(tib$iono.tx.time)), function(tt) ppERKint(tib[tib$iono.tx.time==tt,]))
    names(ppERK_vals) <- sort(unique(tib$iono.tx.time))
    ppERK_vals <- round(ppERK_vals/ppERK_vals[names(ppERK_vals)=="1"],2)

    g <- ggplot(tib, aes(x = avg.pERK, y = iono.tx.time, fill = after_stat(x))) + 
        xlim(-1,1.5) + 
        geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
        scale_fill_viridis_c(name = "avg ppERK", option = "C") +
        labs(title = tit, 
             subtitle=paste(dc,"M ionomycin")) + 
        xlab("Mean ppERK/cell (A.U.)") + ylab("Ionomycin Tx time (min)") + 
        annotate("text", x=1.25, y=1:(length(ppERK_vals))+0.25, label=ppERK_vals, size=2.5) +

        annotate("text", x=1.3, y=1.5, label= paste("n =",nrow(tib[tib$iono.tx.time==1,])), size=2.5) + 
        annotate("text", x=1.3, y=2.5, label= paste("n =",nrow(tib[tib$iono.tx.time==5,])), size=2.5) + 
        annotate("text", x=1.3, y=3.5, label= paste("n =",nrow(tib[tib$iono.tx.time==10,])), size=2.5) + 
        annotate("text", x=1.3, y=4.5, label= paste("n =",nrow(tib[tib$iono.tx.time==15,])), size=2.5) + 
        annotate("text", x=1.3, y=5.5, label= paste("n =",nrow(tib[tib$iono.tx.time==20,])), size=2.5) + 
        
        annotate("text", x=1.2, y=1.75, label= 
                     paste("%pos =",n_vals[n_vals$tx.time==1,]$propos), size=2.5) + 
        annotate("text", x=1.2, y=2.75, label= 
                     paste("%pos =",n_vals[n_vals$tx.time==5,]$propos), size=2.5) + 
        annotate("text", x=1.2, y=3.75, label= 
                     paste("%pos =",n_vals[n_vals$tx.time==10,]$propos), size=2.5) + 
        annotate("text", x=1.2, y=4.75, label= 
                     paste("%pos =",n_vals[n_vals$tx.time==15,]$propos), size=2.5) + 
        annotate("text", x=1.2, y=5.75, label= 
                     paste("%pos =",n_vals[n_vals$tx.time==20,]$propos), size=2.5)
    
    g <- g + theme_ridges()
    g <- g + scale_y_discrete(expand = expand_scale(mult = c(0.15, .75)))

    return(g)
}

```


```{r fig.height=4, fig.width=4, message=FALSE, warning=FALSE}
dc <- 2.5e-6
dtp <- dat[dat$drug1.conc==dc & dat$BRAFi.tx.time=="3 days",c("well","positive.w2","iono.tx.time","avg.pERK")]
dtp <- tibble(dtp)
plotRidges(dtp,dc)
# ggsave("~/Desktop/tol_2.5uMiono_time_ppERK.pdf", width=4, height=4, units="in")
```

```{r fig.height=4, fig.width=4, message=FALSE, warning=FALSE}
dc <- 2.5e-6
dtp <- dat[dat$drug1.conc==dc & dat$BRAFi.tx.time=="60 min",c("well","positive.w2","iono.tx.time","avg.pERK")]
dtp <- tibble(dtp)
plotRidges(dtp,dc,tit="Drug-sensitive cells")
# ggsave("~/Desktop/sens_2.5uMiono_time_ppERK.pdf", width=4, height=4, units="in")
```


```{r fig.height=4, fig.width=4, message=FALSE, warning=FALSE}
dc <- 0
dtp <- dat[dat$drug1.conc==dc & dat$BRAFi.tx.time=="3 days",c("well","positive.w2","iono.tx.time","avg.pERK")]
dtp <- tibble(dtp)
plotRidges(dtp,dc)
# ggsave("~/Desktop/tol_0uMiono_time_ppERK.pdf", width=4, height=4, units="in")
```



