---
title: "Effects of P2Xi on spontaneous spiking"
output: html_notebook
date: 2024-02-19
author: "Darren Tyson & Philip Stauffer"
---

# Data from 20240125
P2Xi on spontaneous Ca2+ spikes in cells treated with BRAFi for 3 days  

## Identifying individual cells (segmented objects) exhibiting spike patterns in calbryte traces
Copying code from "Calbryte trace analysis 20240125 97607PS15.Rmd"

### 2024-01-25 data 97607PS15
```{r}
source("traceAnalysisFxns.r")
```

```{r Load preprocessed data}
intvals <- read.csv("https://www.dropbox.com/scl/fi/nggvx3nph9u68kzwwxfw2/20240125_IXpulsing.P2Xi_ppERK_3dBRAFi_cell_intensities.csv?rlkey=79pr576yc5pn0fbq56h21m83k&dl=1")

wells <- sort(unique(sapply(strsplit(intvals$reg_id, "_"), "[[", 1)))


platemap <- read.csv("../data/platemap_20240125_IXpulsing.P2Xi_ppERK_3dBRAFi.tsv", sep="\t")

platemap <- platemap[platemap$well %in% wells,]

int_well_mean_norm <- lapply(wells, function(w) {
    dff <- intvals[grepl(w,intvals$reg_id),]
    getMeanInt(dff$intensity_mean, unique(dff$reg_id), unique(dff$im_idx))
})

intvals$int_well_mean_norm <- do.call(c, int_well_mean_norm)

rois <- unique(intvals$reg_id)
```

### Number of objects per well
```{r}
sapply(wells, function(w) length(unique(intvals[grepl(w,intvals$reg_id),"reg_id"])))
```


### Analysis of manual quantification
Performed by Philip.  

First-pass analysis will only consider spikes vs no spikes.  
```{r}
f <- list.files("../data/20240125_manualspikedetection")
f <- f[grep("PS.csv", f, fixed=TRUE)]

a <- do.call(rbind, lapply(f, function(x) 
    read.csv(file.path("../data/20240125_manualspikedetection",x))[,c("reg_id", "spikes", "morethan5", "longspikes", "many_longspikes")]))

rois <- a$reg_id
a$well <- sapply(strsplit(a$reg_id,"_"), "[[", 1)
a$cond <- paste(platemap[match(a$well, platemap$well),"drug1"],
                signif(platemap[match(a$well, platemap$well),"drug1.conc"], 1), sep="_")

# exclude data using drugs >1 µM
# a <- a[!grepl("e-06",a$cond),]

all_no_spike_ids <- a$reg_id[a$spikes==0]

wells <- unique(a$well)
conditions <- unique(a$cond)
conditions <- gsub("_6e-06", ", 6 µM", conditions, fixed=TRUE)
conditions <- gsub("_4e-07", ", 400 nM", conditions, fixed=TRUE)
conditions <- gsub("_1e-07", ", 100 nM", conditions, fixed=TRUE)
conditions <- gsub("_0", ", 0 nM", conditions, fixed=TRUE)
platemap[platemap$well %in% wells,]
```

```{r}
set.seed(13)
prop_spiking <- do.call(c, lapply(wells, function(w) {
    n <- 50
    samp <- lapply(1:100, function(i) sample(rois[grepl(w,rois)], n, replace=TRUE))
    sapply(samp, function(x) length(which(!x %in% all_no_spike_ids))/n)
}))

prop_spiking_df <- data.frame(well=rep(wells, each=100), 
                              condition=rep(conditions, each=100),
                              prop_spiking=prop_spiking)
prop_spiking_df$drug <- sapply(strsplit(prop_spiking_df$condition, ","), "[[", 1)
prop_spiking_df$drug_conc <- sapply(strsplit(prop_spiking_df$condition, ", "), "[[", 2)
prop_spiking_df$drug_conc <- factor(prop_spiking_df$drug_conc, levels=c("0 nM", "100 nM", "400 nM", "6 µM"))
```



```{r fig.height=2, fig.width=4}
p <- ggplot(data=subset(prop_spiking_df, !grepl("a74",condition)  & !grepl("aa",condition)), aes(x=drug_conc, y=prop_spiking, fill=1)) + 
    geom_violin(width=.75) + 
    geom_boxplot(width=0.1, color="grey", alpha=0.5) + 
    coord_flip() +
    theme(legend.position="none", 
          axis.text.x = element_text(size=12),
          axis.text.y = element_text(size=12),
          axis.title.x = element_text(size=16, face="bold"),
          axis.title.y = element_text(size=16, face="bold")) + 
    # ylim(c(0,0.75)) +
    ylab("Proportion spiking") + 
    xlab("")
p + facet_wrap(vars(drug))


# ggsave("~/Desktop/P2Xi_spiking.pdf", width=4, height=2, units="in")
```

### Calculate percent inhibition of AZD
```{r}
b <- subset(prop_spiking_df, !grepl("a74",condition)  & !grepl("aa",condition))
z <- b[b$drug=="azd9056",]
z <- z[z$drug_conc %in% c("0 nM", "100 nM"),]
mean_vals <- aggregate(prop_spiking ~ drug_conc, data=z, FUN=mean)
azd_pct_inhib <- 1 - mean_vals[mean_vals$drug_conc=="100 nM","prop_spiking"] / mean_vals[mean_vals$drug_conc=="0 nM","prop_spiking"]
azd_pct_inhib
```
### Calculate percent inhibition of isoPPADS
```{r}
y <- b[b$drug=="isoPPADS",]
y <- y[y$drug_conc %in% c("0 nM", "6 µM"),]
mean_vals <- aggregate(prop_spiking ~ drug_conc, data=y, FUN=mean)
ppads_pct_inhib <- 1 - mean_vals[mean_vals$drug_conc=="6 µM","prop_spiking"] / mean_vals[mean_vals$drug_conc=="0 nM","prop_spiking"]
ppads_pct_inhib
```