---
title: "R Notebook"
output: html_notebook
date: 2024-02-23
author: "Philip Stauffer and Darren Tyson"
---

## Copied from 20240102_analysis_PES_ridges.Rmd

# ATP stim: 20240102_ATPstim_3dayPLX_longstim -- FBS free media in the assay
First pass visualizing the results of the experiment (details can be found on elab).
See `~/Melanoma/Ca2Mobilization/20240102_ATPstim_3dayPLX` directory for thunor .tsv platemap and for barcode information.

pERK staining procedure and imaging details can be found in the elab entry titled 20231116_ATPstim

## Data processing details
MetaExpress multiwavelength cell scoring analysis performed with the settings saved in:
"VQU_PES_Ca2+ mobilization"


```{r Setup}
library(tidyr)
library(ggplot2)
library(ggridges)
library(dplyr)
library(viridis)
```

### Function to quantify ppERK using both proportion of positive cells and the avg intensity
Using both proportion of cells staining positively for ppERK and the integrated values of the positive cells.  

```{r}
ppERKint <- function(dat, signif=4) {
    cell_count <- nrow(dat)
    pos_cells <- nrow(dat[dat$positive.w2==1,])
    val <- pos_cells/cell_count*sum(dat[dat$positive.w2==1,]$avg.pERK)
    val <- signif(val,signif)
    return(val)
}
```

## Load data
```{r}
fn <- "https://www.dropbox.com/scl/fi/xgjbo711hr5mru5axfhh3/20240102_singlecell_all_ATP.csv?rlkey=fzbzg5h0g25ponw1igw5kizdu&dl=1"
d <- read.csv(fn)
```

# QC
Removing cells with total area > 4500 and < 100 $pixels^2$  
Removing objects with integrated intensity > 5e6 because they tend to be bubbles, precipitated antibody, or debris.
It was found that setting minimum area to above 200 pixels^2 predominantly filtered out data from the Ca2+ free control group. Because of this bias in filtering, minimum area was set to > 100 pixels. 

### Filtering (QC)
NOTE: DRT increased filter for total.area to 100 pixels^2
```{r}
o_nrow <- nrow(d)
d <- d[d$total.area < 4500 & d$total.area > 100 & d$w2.stain.integr.intens <5e6,]
message(cat("Current nrow = ",nrow(d),"\nOriginal nrow = ",o_nrow))
message(cat("QC removed",signif(100-(nrow(d)/o_nrow*100),2),"% of cells"))
```


### Annotation
* Columns 17 and 18: untreated control; all others: 8 µM PLX-4720 for 3 days 
** column 17 had serum free media added; column 18 had serum containing media added
* When `drug1== "persistent"`, that is the no ATM stim control (within column 17) for a background reference
* When `drug1== "veh"`, that is when there is ATP added without inhibitor present; this is the baseline response for each drug at each time point.
* When `drug1== "control"`, that is the drug naive control (column 18) with serum free media added
* When `drug1== ""` & `drug2== ""`, that is the antibody control


```{r}
d$drug1[d$drug1== ""] <- "abctrl"
ab_ctrl_wells <- unique(d[d$drug1=="abctrl",]$well)
untx_ctrl_wells <- unique(d[d$drug1=="control",]$well)
ctrl_wells <- c(ab_ctrl_wells,untx_ctrl_wells)

d[is.na(d$drug1.conc),"drug1.conc"] <- -1


tx <- tibble(d[!d$well %in% ctrl_wells, ])
tx <- tx[order(tx$BRAFi.tx.time),]

rownames(tx) <- NULL

ctrls <- d[d$well %in% ctrl_wells, ]

tx$drug1.conc <- signif(tx$drug1.conc, digits= 2)
tx$drug2[tx$drug1== "persistent"] <- 0
```


```{r message=FALSE, warning=FALSE}
ggplot(d[d$well %in% ctrl_wells, ], aes(x = avg.pERK, y = drug1, fill = after_stat(x))) + 
    xlim(-1,1.5) + 
    geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
    scale_fill_viridis_c(name = "avg ppERK", option = "C") +
    labs(title = 'Controls', 
         subtitle="") + 
    xlab("Mean ppERK/cell (A.U.)") + ylab("Control condition")

ggplot(tx[tx$drug1 %in% c("persistent", "veh"),], aes(x = avg.pERK, y = drug1, fill = after_stat(x))) + 
    xlim(-1,1.5) + 
    geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
    scale_fill_viridis_c(name = "avg ppERK", option = "C") +
    labs(title = 'ATP stim vs no ATP stim', 
         subtitle="") + 
    xlab("Mean ppERK/cell (A.U.)") + ylab("Control condition")

```


### Determine percent positive for each condition using bootstrapping
```{r}
f <- tx[,c("cell.id", "plate.id", "well", "BRAFi.tx.time", "atp.tx.time", "drug1", "drug1.conc", "total.area", "positive.w2","w2.stain.integr.intens")]

f <- f[f$atp.tx.time %in% c(0,1,30),]
f <- f[order(f$plate.id, f$well),]
rownames(f) <- NULL

f$uid <- paste(f$plate.id, f$well, sep="_")
uids <- unique(f$uid)
f$condition <- paste(f$plate.id, f$well, f$atp.tx.time, f$drug1, f$drug1.conc, sep="_")
conditions <- f[!duplicated(f$uid),]$condition
conditions_plates <- sapply(strsplit(conditions, "_"), "[[", 1)
conditions_atp_time <- sapply(strsplit(conditions, "_"), "[[", 3)

no_ppERK_ids <- f[f$positive.w2==0,]$cell.id
```

### Drugs on each plate
```{r}
unique(f[f$plate.id=="142708",]$drug1)
```

```{r}
unique(f[f$plate.id=="142709",]$drug1)
```


```{r}

set.seed(13)
n <- 1000

prop_pos <- do.call(c, lapply(uids, function(id) {
    cellids <- f[f$uid==id,]$cell.id
    cid_samples <- lapply(1:10, function(i) sample(cellids, n))
    unlist(lapply(cid_samples, function(cids) length(which(f[f$cell.id %in% cids,]$positive.w2==1))/nrow(f[f$cell.id %in% cids,])))
}))

prop_pos_df <- data.frame(uid=rep(uids, each=10), 
                              condition=rep(conditions, each=10),
                              prop_pos=prop_pos)
```

```{r}
prop_pos_df$cond2 <- gsub("_[A-P][0-2][0-9]_","_",prop_pos_df$condition)
mean_props <- sapply(unique(prop_pos_df$cond2), function(co) mean(prop_pos_df[prop_pos_df$cond2==co,]$prop_pos))

veh_mean_props <- mean_props[grep("veh", names(mean_props))]
veh_mean_prop_range <- sapply(142708:142709, function(x) 
    veh_mean_props[grep(x,names(veh_mean_props))][1] - 
      veh_mean_props[grep(x,names(veh_mean_props))][2])

b <- prop_pos_df[(grepl("_30_",prop_pos_df$condition) & 
                    grepl(9.8e-08, prop_pos_df$condition)) | grepl("veh", prop_pos_df$condition),]
b$drug1 <- sapply(strsplit(b$condition, "_"), "[[", 4)
b$drug1.conc <- as.numeric(sapply(strsplit(b$condition, "_"), "[[", 5))
drugs <- unique(b$drug1)

b$prop_pos_diff <- b$prop_pos - mean_props[gsub("_30_","_1_",b$cond2)]
b$prop_pos_ratio <- ifelse(grepl("142708",b$uid), b$prop_pos_diff/veh_mean_prop_range[1], b$prop_pos_diff/veh_mean_prop_range[2])

```

```{r all drugs, fig.height=2, fig.width=3}
dtp <- subset(b, !grepl("_1_", condition) & drug1 %in% c("a740003","aacba","azd9056","isoPPADS", "bx430", "veh"))
dtp$drug1 <- factor(dtp$drug1, levels=c("isoPPADS","azd9056","a740003","aacba", "bx430", "veh"))
mycolors <- viridis::magma(7, direction=-1)[-1]

ggplot(data=dtp, aes(x= factor(drug1), y= 100 * (1 - prop_pos_ratio), fill=factor(drug1))) +
  # geom_boxplot() +
  geom_violin() +
  coord_flip() +
  scale_fill_manual(values=mycolors) +  # , guide = guide_legend(reverse = TRUE)
  xlab("") +
  ylab("% inhib, ATP-induced ppERK+") +
  theme(legend.position = "none")

# ggsave("~/Desktop/P2Xi_ATP_dose-dependent_ppERK_drug-tol.pdf", width=3, height=2, units="in", device="pdf")
```

```{r}
m <- lm(prop_pos_ratio ~ drug1, data=dtp)
m_anova <- aov(m)
m_tukey <- TukeyHSD(m_anova)
m_tukey$drug1[grep("veh",rownames(m_tukey$drug1)),]
```

